---
title: "Vizathon 2015 - Hidden Hunger"
author: "Angel Mak"
date: "May 30, 2015"
output: html_document
---

## Summary  
Using the Malawi household data, this exploratory analysis found a positive correlation between wealth and education(b6). The education level was recoded assuming that "Std" were primary level of education. It was found that diversity of diet is positively correlated with education level. However, there was an unexpected negative correlation of number of deficient micronutrients with education level. This relationship was more profound when households that were not deficient in calories and proteins were removed.

Data source:  
https://github.com/harvestchoice/nutrition_indicators

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
fp.mwi.nutr <- "/Users/angelmak/Dropbox/git/nutrition_indicators/Malawi/mwi_nutrition.csv"
t.m.nutr <- read.csv(fp.mwi.nutr)

# variables indicating deficiency in nutrients, in the required level
req.names <- names(t.m.nutr)[grep("req_hh",names(t.m.nutr))]
# variables indicating deficiency in micro-nutrients(exclude calories and proteins), in the required level
req.names.micro <- req.names[3:14]
req.names.micro
for (i in 1:nrow(t.m.nutr)){
  t.m.nutr[i,"defCountReq"] <- sum(t.m.nutr[i,req.names] == "Yes" )
}
for (i in 1:nrow(t.m.nutr)){
  t.m.nutr[i,"defCountReqMicro"] <- sum(t.m.nutr[i,req.names.micro] == "Yes")
}
```

## A positive relationship of household wealth and education level (b6)

```{r, message=FALSE, warning=FALSE}
#Education -> wealth
options("scipen"=100, "digits"=4) #Help print numeric text into numbers
plot <- ggplot() + theme(panel.background = element_rect(fill=0,colour="black"), axis.ticks=element_blank(), panel.grid = element_blank()) 
plot <- plot + theme(axis.text.x = element_text(angle=45, hjust =1))
plot <- plot + geom_boxplot(data=t.m.nutr,aes(t.m.nutr$b6,t.m.nutr$wealth))
plot <- plot + labs(title=paste("Wealth and Education Level",sep="")) #+ theme(plot.title = element_text(size = rel(2), colour = "blue"))
plot <- plot + xlab("Education Level (b6)") # x-axis title
plot <- plot + ylab("Wealth") # y-axis title
plot
```

## Recoding education level

The education level was recoded to show a clearer trend:  
1=primary (No school and all "Std" levels)  
2=secondary ("Form" levels)  
3=tertiary  

```{r, message=FALSE, warning=FALSE}
#Recode education level
#1=primary
#2=secondary
#3=tertiary
rowi.2 <- grep("Form",t.m.nutr$b6)
rowi.1 <- grep("Std|No",t.m.nutr$b6)
rowi.3 <- grep("College|Diploma",t.m.nutr$b6)
t.m.nutr$edLvl[rowi.1] <- 1
t.m.nutr$edLvl[rowi.2] <- 2
t.m.nutr$edLvl[rowi.3] <- 3

#Using the new variable against wealth
plot <- ggplot() + theme(panel.background = element_rect(fill=0,colour="black"), axis.ticks=element_blank(), panel.grid = element_blank()) 
plot <- plot + theme(axis.text.x = element_text(angle=45, hjust =1))
plot <- plot + geom_boxplot(data=t.m.nutr,aes(factor(t.m.nutr$edLvl),t.m.nutr$wealth))
plot <- plot + xlab("Education Level (recoded)") # x-axis title
plot <- plot + ylab("Wealth") # y-axis title
plot

table(t.m.nutr$edLvl,t.m.nutr$def_kcal_req_hh)
table(t.m.nutr$edLvl,t.m.nutr$def_prot_req_hh)
```

## A positive relationship between education level and diversity of diet (Shannon index and Simpson index)

```{r, message=FALSE, warning=FALSE}
# Diversity of food
plot <- ggplot() + theme(panel.background = element_rect(fill=0,colour="black"), axis.ticks=element_blank(), panel.grid = element_blank()) 
plot <- plot + theme(axis.text.x = element_text(angle=45, hjust =1))
plot <- plot + xlab("Education Level (recoded)") # x-axis title
plot <- plot + ylab("Diversity of diet (Shannon index)") # y-axis title
plot <- plot + geom_boxplot(data=t.m.nutr,aes(factor(t.m.nutr$edLvl),t.m.nutr$shannon_exp))
plot

plot <- ggplot() + theme(panel.background = element_rect(fill=0,colour="black"), axis.ticks=element_blank(), panel.grid = element_blank()) 
plot <- plot + theme(axis.text.x = element_text(angle=45, hjust =1))
plot <- plot + geom_boxplot(data=t.m.nutr,aes(factor(t.m.nutr$edLvl),t.m.nutr$simpson_exp))
plot <- plot + xlab("Education Level (recoded)") # x-axis title
plot <- plot + ylab("Diversity of diet (Simpson index)") # y-axis title
plot
```

## A negative relationship was noticed between education level and the number of deficient micro-nutrients  

The higher the education level of the households, the more the number of micronutrients they are missing.


```{r, message=FALSE, warning=FALSE}
t.m.nutr.lite <- t.m.nutr[which(t.m.nutr$def_kcal_rec_hh == "No" & t.m.nutr$def_prot_rec_hh == "No"),]
#t.m.nutr.lite[1:10,257:284]


# Deficiency
plot <- ggplot() + theme(panel.background = element_rect(fill=0,colour="black"), axis.ticks=element_blank(), panel.grid = element_blank()) 
plot <- plot + theme(axis.text.x = element_text(angle=45, hjust =1))
plot <- plot + geom_boxplot(data=t.m.nutr,aes(factor(t.m.nutr$edLvl),t.m.nutr$defCountReqMicro))
plot <- plot + xlab("Education Level") # x-axis title
plot <- plot + ylab("Number of deficient micronutrients") # y-axis title
plot
```

  
The negative relationship becomes more prominent in the tertiary education level after removing households that were not deficient in calories and proteins  


```{r, message=FALSE, warning=FALSE}
# Only on people who are not deficient in calory and protein
plot <- ggplot() + theme(panel.background = element_rect(fill=0,colour="black"), axis.ticks=element_blank(), panel.grid = element_blank()) 
plot <- plot + theme(axis.text.x = element_text(angle=45, hjust =1))
plot <- plot + geom_boxplot(data=t.m.nutr.lite,aes(factor(t.m.nutr.lite$edLvl),t.m.nutr.lite$defCountReqMicro))
plot <- plot + labs(title=paste("N=723/1124 (Household not deficient in kcal and protein)",sep="")) #+ theme(plot.title = element_text(size = rel(2), colour = "blue"))
plot <- plot + xlab("Education Level") # x-axis title
plot <- plot + ylab("Number of deficient micronutrients") # y-axis title
plot
```

Total number of household:
```{r}
nrow(t.m.nutr)
```

Number of household after removing those that were not deficient in carlories and proteins
```{r}
nrow(t.m.nutr.lite)
```

Note that the number of households is small for those with Tertiary level of education

```{r}
summary(factor(t.m.nutr.lite$edLvl))
```
